Inducing Domain Specific Languages for Bayesian Program Learning

### 2018-02-06 paper.tex
miscellaneous improvements

\begin{abstract}
This document provides a basic paper template and submission guidelines.
Abstracts must be a single paragraph, ideally between 4--6 sentences long.
Gross violations will trigger corrections at the camera-ready phase.
\end{abstract}

@@

Our algorithm is best understood as inference in a hierarchical Bayesian model (Figure~\ref{graphicalModel}(a): black lines):
A DSL $\mathcal{D}$, equipped with a real-valued weight vector $\theta$,
give a distribution over programs
from which a latent program is sampled,
which then determines the agent's observations.
Our goal will be both to infer the programs that best explain our observations,
and at the same time induce a DSL
that distills the commonalities across all of the programs that solve the tasks.
Inducing the DSL means learning the symbolic structure of a latent programming language,
and so we are confronted with two distinct structure learning problems:
the structure of the programs and the structure of the DSL.

Inference in the generative model in Figure~\ref{graphicalModel}(a) is
difficult because of the vastness of the combinatorial space of all
programs. To make inference tractable our model learns a bottom-up
\emph{recognition model} alongside the top-down generative model
(Figure~\ref{graphicalModel}b).  The recognition model $q$ is a neural
network that regresses from observations to a distribution over
programs.  The neural network $q$ implements an amortized
inference/compiled inference
scheme~\cite{le2016inference,ritchie2016deep} for the generative
model.

@@

The recognition model $q$ is a neural network that predicts,
for each task $x\in X$, a weight vector $q(x) = \theta^{(x)}\in \mathbb{R}^{|\mathcal{D}| + 1}$.
Together with the DSL, this defines a distribution over programs,
$\probability[p|\mathcal{D},\theta = q(x)]$.
We abbreviate this distribution as $q(p|x)$.
The crucial aspect of this framing is that the neural network
leverages the structure of the learned DSL,
so it is \emph{not} responsible for
generating programs wholesale.

@@

@@

Exact maximization of $\lowerBound$ (Eq.\ref{lowerBound}) w.r.t. $(\mathcal{D},\theta)$
is intractable,
so we take a more heuristic approach.
The strategy is to search locally through the space of DSLs,
proposing small local changes to $\mathcal{D}$ until $\lowerBound$ fails to increase.
The search moves work by introducing new
$\lambda$-expressions into the DSL.
We propose these new expressions by extracting subexpressions from
programs already in the frontier.
These extracted subexpressions
are fragments of the original programs, and can introduce new variables (Figure~\ref{fragmentExample}),
which then become new functions in the DSL.

%% Our model of subexpressions is loosely related to Fragment
%% Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing},
%% but context-sensitive.

The idea of storing and reusing
fragments of expressions comes from Fragment Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing}.
Somehow I need to clarify that this is not like an adapter grammar, and that previous work has done adapter grammar like things... In fact this might be a baseline, because with the right parameter settings our algorithm can revert to an adapter grammar like phase.

@@

\begin{figure*}
  \begin{tabular}{lll}
    \toprule
    Domain&Example programs in frontiers&Example proposed subexpression\\\midrule
    Boolean circuits&
    \begin{tabular}{l}
      \texttt{(lambda x (nand x x))}\\
      \texttt{(lambda x (lambda y (nand x (nand y y))}
    \end{tabular}&
    \texttt{(nand z z)}\\\midrule
    String editing&
    \begin{tabular}{l}
      \texttt{(lambda s (+ ',' (index 0 (split ',' s))))}\\
      \texttt{(lambda s (index 0 (split ' ' s)))}
    \end{tabular}&
    \texttt{(index 0 (split z s))}\\\bottomrule
  \end{tabular}
  \caption{The DSL induction algorithm works by proposing subexpressions of programs to add to the DSL.
    These subexpressions are taken from programs in the frontiers (middle column), and can introduce new variables (\texttt{z} in the right column).}
  \label{fragmentExample}
\end{figure*}

@@

\begin{algorithm}[tb]
   \caption{The \system Algorithm}
   \label{mainAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Initial DSL $\mathcal{D}$, set of tasks $X$, iterations $I$
     \STATE \textbf{Hyperparameters:} Maximum frontier size $F$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$, recognition model $q(\cdot)$
     \STATE Initialize $\theta\gets \text{uniform}$ %, $q_0(\cdot ) = \theta_0$
     \FOR{$i=1$ {\bfseries to} $I$}
%     \FOR{$x:\tau\in X$}
     \STATE  $\mathcal{F}^{(\theta)}_x\gets \{z| z\in \text{enumerate}(\mathcal{D},\theta,F)\text{ if }\probability[x|z] > 0\}$
 %    \ENDFOR
     \STATE $q_i\gets \text{train recognition model to maximize }\mathcal{L}_{\text{RM}}$
     \STATE  $\mathcal{F}^{(q)}_x\gets\{z|z\in \text{enumerate}(\mathcal{D},q(x),F)\text{ if }\probability[x|z] > 0\}$
     \STATE $\mathcal{D},\theta\gets $induceDSL$(\{\mathcal{F}^{(\theta)}_x\cup\mathcal{F}^{(q)}_x\}_{x\in X})$
     %% \STATE Define $Q_x(z) \propto \begin{cases}
     %%   \probability[x|z]\probability[z|\mathcal{D}_i,\theta_i]&x\in \mathcal{F}_x\\
     %%   0&x\not \in \mathcal{F}_x
     %% \end{cases}
      \ENDFOR
 \STATE \textbf{return} $\mathcal{D},\theta,q$
\end{algorithmic}
\end{algorithm}

@@

\subsection{Boolean circuits}
As a toy warm-up domain, we consider the problem of learning
to build the Boolean circuits out of logic gates.
Now, if we equipped our agent with a full repertoire of
circuit primitives -- XOR gates, multiplexers, etc. -- this problem would be easy,
and we could use sophisticated symbolic search algorithms like SMT solvers to synthesize circuits.
Instead, we gave \system only a NAND gate,
and tasked it with building a set of 500 random circuits made out of
AND, NOT, and OR gates.
This experimental setup was introduced in the original EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}.
For this domain our recognition model is a simple multilayer perceptron
whose inputs are the input/output examples.
Figure~\ref{circuitLearningCurve} shows our algorithm learning to build circuits; Table~\ref{circuitExperimentTable}

@@

\begin{table}
\begin{tabular}{ll}
  \toprule
  Primitives&\texttt{nand}

\\\midrule
  Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  \midrule
  Likelihood $\probability[x|p]$& $\prod_n \indicator[p(i_n) = o_n]$
  \\\midrule
  \begin{tabular}{l}
    Subset of \\Learned DSL
    \end{tabular}&

  \\\bottomrule
\end{tabular}
\caption{}\label{circuitExperimentTable}
\end{table}

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Table~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom:
for example, it represents the linear function $-x+2$ with the program 
\texttt{(* (* (+ x }$\mathcal{R}$\texttt{) }$\mathcal{R}$\texttt{) x)}, which has two continuous degrees of freedom -- while representing quartic functions using the invented DSL primitive $f_6$ in Table~\ref{regressionDSL}.
This falls out of our Bayesian treatment of the program learning problem -- both the bias towards shorter programs (more likely under $\mathcal{D},\theta$) and the likelihood's model BIC penalty.
\begin{table}
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
    $f_0 = $\texttt{(+ }$\mathcal{R}$\texttt{)}\\
    $f_1($\texttt{x}$) =$ \texttt{(* (* (+ x }$\mathcal{R}$\texttt{) }$\mathcal{R}$\texttt{) x)} \\
      $f_2($\texttt{x}$) =$ \texttt{(+ }$\mathcal{R}$\texttt{ ($f_1$ x))} \\
      $f_3($\texttt{x}$) = $\texttt{(* x (}$f_2$\texttt{ x))}\\
  $f_4($\texttt{x}$) = $\texttt{(}$f_0$\texttt{ (}$f_3$\texttt{ x)))}\\
  $f_5($\texttt{x}$) =$ \texttt{(* x (}$f_4$\texttt{ x))}\\
  $f_6($\texttt{x}$) = $\texttt{(}$f_0$\texttt{ (}$f_4$\texttt{ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some induced DSL primitives for symbolic regression. System starts out with addition, multiplication, and real numbers, and learns functions for building 4-th order polynomials.}\label{regressionDSL}
\end{table}

\begin{figure}\centering
  \includegraphics[width = 7cm]{figures/symbolicRegression.png}
  \caption{DSL learning curve for symbolic regression.}\label{regressionLearningCurve}
\end{figure}
\subsection{String editing}\label{textSection}

\begin{tabular}{ll}
  \toprule
  Primitives&\begin{tabular}{l}
    \texttt{0}, \texttt{+1},\texttt{-1}\\
    \texttt{split}, \texttt{join}, \texttt{map}, \texttt{length}, \texttt{slice}\\
    \texttt{append}, \texttt{chr->str}\\
    \texttt{lowercase}, \texttt{uppercase}, \texttt{capitalize}\\
    \texttt{""},    \texttt{","},     \texttt{" "},    \texttt{"<"}, \texttt{">"}
    \end{tabular}

\\\midrule
  Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  \midrule
  Likelihood $\probability[x|p]$& $\prod_n \indicator[p(i_n) = o_n]$
  \\\midrule
  \begin{tabular}{l}
    Subset of \\Learned DSL
    \end{tabular}&

  \\\bottomrule
  \end{tabular}
\begin{figure}

>>>>

\begin{abstract}

\end{abstract}

@@

Our algorithm is best understood as inference in a hierarchical Bayesian model (Figure~\ref{graphicalModel}(a): black lines):
A DSL $\mathcal{D}$, equipped with a real-valued weight vector $\theta$,
defines a distribution over programs.
A program is sampled from $(\mathcal{D},\theta)$,
and then the program is executed to determines the agent's observations.
Our goal will be both to infer the programs that best explain our observations,
and at the same time induce a DSL
that distills the commonalities across all of the programs that solve the tasks.
%% Inducing the DSL means learning the symbolic structure of a latent programming language,
%% and so we are confronted with two distinct structure learning problems:
%% the structure of the programs and the structure of the DSL.

Inference in the generative model in Figure~\ref{graphicalModel}(a) is
difficult because the programs are unobserved, and so we must solve a
difficult search problem to find any programs at all. To make
inference tractable our model learns a bottom-up \emph{recognition
  model} alongside the top-down generative model
(Figure~\ref{graphicalModel}b).  The recognition model $q$ is a neural
network that regresses from observations to a distribution over
programs.  The neural network $q$ implements an amortized
inference/compiled inference
scheme~\cite{le2016inference,ritchie2016deep} for the generative
model.

@@

The recognition model $q$ is a neural network that predicts,
for each task $x\in X$, a weight vector $q(x) = \theta^{(x)}\in \mathbb{R}^{|\mathcal{D}| + 1}$.
Together with the DSL, this defines a distribution over programs,
$\probability[p|\mathcal{D},\theta = q(x)]$.
We abbreviate this distribution as $q(p|x)$.
The crucial aspect of this framing is that the neural network
leverages the structure of the learned DSL,
so it is \emph{not} responsible for
generating programs wholesale.
This taps into the intuition that programming is primarily a top-down activity:
as human programmers, we first decide what kind of
programming constructs we might need to use,
and then we figure out how to assemble them in to the desired program.

@@

The neural architecture of $q$ depends upon the domain;
for domains with sequential structure like strings and lists,
we use an RNN,
and for domains with a fixed-dimensionality inputs/outputs we use a multilayer perceptron.

@@

Recall from Section~\ref{mathematicalFraming} that we want the DSL maximizing $\int \lowerBound\;\mathrm{d}\theta$.
We replace this marginal with an AIC approximation, giving the following objective for DSL induction:
\begin{align}
\nonumber  \log \probability[\mathcal{D}] + \argmax_{\theta}& \sum_{x\in X}\log \sum_{p\in \mathcal{F}_x}\probability[x|p]\probability[p|\mathcal{D},\theta]\\
&+  \log P(\theta|\mathcal{D}) - ||\theta||_0 \label{AIC}
  \end{align}

We induce a DSL by searching locally through the space of DSLs,
proposing small local changes to $\mathcal{D}$ until Eq.~\ref{AIC} fails to increase.
The search moves work by introducing new
$\lambda$-expressions into the DSL.
We propose these new expressions by extracting subexpressions from
programs already in the frontier.
These extracted subexpressions
are fragments of the original programs, and can introduce new variables (Figure~\ref{fragmentExample}),
which then become new functions in the DSL.
The idea of storing and reusing
fragments of expressions comes from Fragment Grammars~\cite{tim} and Tree-Substitution Grammars~\cite{cohn2010inducing}.

@@

\begin{figure*}
\centering  \begin{tabular}{lll}
    \toprule
    Domain&Example programs in frontiers&Example proposed subexpression\\\midrule
    Circuits&
    \begin{tabular}{l}
      \texttt{(lambda x (nand x x))}\\
      \texttt{(lambda x (lambda y (nand x (nand y y))}
    \end{tabular}&
    \texttt{(nand z z)}\\\midrule
    Strings&
    \begin{tabular}{l}
      \texttt{(lambda s (+ ',' (index 0 (split ',' s))))}\\
      \texttt{(lambda s (index 0 (split ' ' s)))}
    \end{tabular}&
    \texttt{(index 0 (split z s))}\\\bottomrule
  \end{tabular}
  \caption{The DSL induction algorithm works by proposing subexpressions of programs to add to the DSL.
    These subexpressions are taken from programs in the frontiers (middle column), and can introduce new variables (\texttt{z} and \texttt{s} in the right column).}
  \label{fragmentExample}
\end{figure*}

@@

\begin{algorithm}[tb]
   \caption{The \system Algorithm}
   \label{mainAlgorithm}
   \begin{algorithmic}
     \STATE {\bfseries Input:} Initial DSL $\mathcal{D}$, set of tasks $X$, iterations $I$
     \STATE \textbf{Hyperparameters:} Maximum frontier size $F$
     \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$, recognition model $q(\cdot)$
     \STATE Initialize $\theta\gets \text{uniform}$ %, $q_0(\cdot ) = \theta_0$
     \FOR{$i=1$ {\bfseries to} $I$}
%     \FOR{$x:\tau\in X$}
     \STATE  $\mathcal{F}^{\theta}_x\gets \{p| p\in \text{enumerate}(\mathcal{D},\theta,F)\text{ if }\probability[x|p] > 0\}$
 %    \ENDFOR
     \STATE $q(\cdot )\gets \text{train recognition model to maximize }\mathcal{L}_{\text{RM}}$
     \STATE  $\mathcal{F}^{q}_x\gets\{p|p\in \text{enumerate}(\mathcal{D},q(x),F)\text{ if }\probability[x|p] > 0\}$
     \STATE $\mathcal{D},\theta\gets $induceDSL$(\{\mathcal{F}^{\theta}_x\cup\mathcal{F}^{q}_x\}_{x\in X})$
     %% \STATE Define $Q_x(z) \propto \begin{cases}
     %%   \probability[x|z]\probability[z|\mathcal{D}_i,\theta_i]&x\in \mathcal{F}_x\\
     %%   0&x\not \in \mathcal{F}_x
     %% \end{cases}
      \ENDFOR
 \STATE \textbf{return} $\mathcal{D},\theta,q$
\end{algorithmic}
\end{algorithm}

@@

\subsection{Boolean circuits}
As a toy domain, we consider the problem of learning
to build the Boolean circuits out of logic gates.
Now, if we equipped our agent with a full repertoire of
circuit primitives -- XOR gates, multiplexers, etc. -- this problem would be easy,
and we could use sophisticated symbolic search algorithms like SMT solvers to synthesize circuits.
Instead, we gave \system only a NAND gate,
and tasked it with building a set of 500 random circuits made out of
AND, NOT, and OR gates,
as well as 2 or 4 input multiplexers.
This experimental setup is an extension of one introduced in the original EC algorithm~\cite{Dechter:2013:BLV:2540128.2540316}.
For this domain our recognition model is a simple multilayer perceptron
whose inputs are the input/output examples.
Figure~\ref{circuitLearningCurve} shows our algorithm learning to build circuits; Table~\ref{circuitExperimentTable}

@@

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Table~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom:
for example, it represents the linear function $-x+2$ with the program 
\texttt{(* (* (+ x }$\mathcal{R}$\texttt{) }$\mathcal{R}$\texttt{) x)}, which has two continuous degrees of freedom -- while representing quartic functions using the invented DSL primitive $f_6$ in Table~\ref{regressionDSL}.
This falls out of our Bayesian treatment of the program learning problem -- both the bias towards shorter programs (more likely under $\mathcal{D},\theta$) and the likelihood's model BIC penalty.
\begin{table}
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
    $f_0 = $\texttt{(+ }$\mathcal{R}$\texttt{)}\\
    $f_1($\texttt{x}$) =$ \texttt{(* (* (+ x }$\mathcal{R}$\texttt{) }$\mathcal{R}$\texttt{) x)} \\
      $f_2($\texttt{x}$) =$ \texttt{(+ }$\mathcal{R}$\texttt{ ($f_1$ x))} \\
      $f_3($\texttt{x}$) = $\texttt{(* x (}$f_2$\texttt{ x))}\\
  $f_4($\texttt{x}$) = $\texttt{(}$f_0$\texttt{ (}$f_3$\texttt{ x)))}\\
  $f_5($\texttt{x}$) =$ \texttt{(* x (}$f_4$\texttt{ x))}\\
  $f_6($\texttt{x}$) = $\texttt{(}$f_0$\texttt{ (}$f_4$\texttt{ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some induced DSL primitives for symbolic regression. System starts out with addition, multiplication, and real numbers, and learns functions for building 4th order polynomials.}\label{regressionDSL}
\end{table}

\begin{figure}\centering
  \includegraphics[width = 7cm]{figures/symbolicRegression.png}
  \caption{DSL learning curve for symbolic regression.}\label{regressionLearningCurve}
\end{figure}
\subsection{String editing}\label{textSection}
\begin{table}

  \centering
  \begin{tabular}{l}
  \toprule
    \texttt{0}, \texttt{+1},\texttt{-1}\\
    \texttt{split}, \texttt{join}, \texttt{map}, \texttt{length}, \texttt{slice}\\
    \texttt{append}, \texttt{chr->str}, \texttt{trim}\\
    \texttt{lowercase}, \texttt{uppercase}, \texttt{capitalize}\\
    \texttt{""},    \texttt{","},     \texttt{" "},    \texttt{"<"}, \texttt{">"}  
  \\\bottomrule
  \end{tabular}
  \caption{Initial set of primitives provided to the system for learning string editing programs}
  \end{table}
\begin{figure}
